---
title: "Outhitting the Odds"
date: "2024-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### code so far
#### Ben Scartz, Lauren Beaudreau, and Jack Arbuckle


For each hitter:

-   % swings 90+ EV for each pitch (R/L)
-   ZO Swing % for each pitch
-   xBA last 25 PA
-   xBA overall (R/L)
-   PA per game

For each pitcher: 

- % swings 90+ EV for each pitch (R/L)
- % usage of each pitch (R/L)
- xBA last 75 BF 
- BF per start

For each pitcher team: 

- bullpen xBA last 150 BF


Historical Response

- date
- hitter
- starting pitcher
- pitcher team
- hit (1 / 0) or # of hits ?

```{r warning=FALSE, include=FALSE}
library(tidyverse)
library(baseballr)
```
```{r}
#library(tidyverse)
#library(baseballr)
```

```{r}
statcast_2023 <- read.csv('statcast_2023.csv')
statcast_2022 <- read.csv('statcast_2022.csv')
statcast_full <- rbind(statcast_2023, statcast_2022)
```
Prep

```{r}
swings_vector <- c('hit_into_play', 'swinging_strike_blocked', 'swinging_strike', 'foul_tip', 'foul')
hits_vector <- c('single', 'double', 'triple', 'home_run')
at_bats_vector <- c('field_out', 'strikeout', 'grounded_into_double_play', 'fielders_choice', 'fielders_choice_out', 'triple_play', 'strikeout_double_play', 'double_play', 'field_error', 'force_out', hits_vector)
plate_appearances_vector <- c(at_bats_vector, 'sac_fly', 'walk', 'hit_by_pitch', 'sac_bunt', 'sac_bunt_double_play', 'sac_fly_double_play', 'catcher_interf')
swings_vector <- c('hit_into_play', 'swinging_strike_blocked', 'swinging_strike', 'foul_tip', 'foul')

# add strikeouts to xBA and batter / pitcher team
statcast_full <- statcast_full %>%
  mutate(estimated_ba_using_speedangle = if_else(events == 'strikeout', 0, estimated_ba_using_speedangle),
         batter_team = case_when(inning_topbot == 'Top' ~ away_team,
                                 inning_topbot == 'Bot' ~ home_team),
         pitcher_team = case_when(inning_topbot == 'Top' ~ home_team,
                                 inning_topbot == 'Bot' ~ away_team))

```

#### Historical hits response data

```{r}
game_data <- statcast_full %>%
  group_by(game_pk, batter) %>%
  summarize(date = first(game_date),
            hits = sum(events %in% hits_vector),
            batter_team = first(batter_team),
            pitcher_team = first(pitcher_team),
            home_team = first(home_team),
            away_team = first(away_team))

home_starter <- statcast_full %>%
  filter(inning_topbot == 'Top') %>%
  group_by(game_pk) %>%
  summarize(home_starter = first(pitcher))

away_starter <- statcast_full %>%
  filter(inning_topbot == 'Bot') %>%
  group_by(game_pk) %>%
  summarize(away_starter = first(pitcher))

starters <- home_starter %>%
  left_join(away_starter, by = 'game_pk')

game_data <- game_data %>%
  left_join(starters, by = 'game_pk') %>%
  mutate(opp_starter = case_when(
                                  batter_team == away_team ~ home_starter,
                                  batter_team == home_team ~ away_starter
  )) %>%
  select(-home_team, -away_team)


head(game_data)
```


#### Hitters

90+

```{r}
# By pitch & side
by_pitch_90 <- data %>%
  group_by(player_name, p_throws, pitch_name) %>%
  summarize(pitch_90 = sum(launch_speed >= 90, na.rm = TRUE) /
                            sum(description %in% swings_vector, na.rm = TRUE),
            swings = sum(description %in% swings_vector, na.rm = TRUE)) %>%
  mutate(pitch_90 = if_else(swings < 10, NA, pitch_90))

batter_90 <- by_pitch_90 %>%
  mutate(combined_name = paste(pitch_name, p_throws, "90", sep = "_")) %>%
  ungroup() %>%
  select(-p_throws, -swings, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = pitch_90) 


```

xBA

```{r}
batter_xba <- data %>%
  group_by(player_name) %>%
  summarize(xba = sum(estimated_ba_using_speedangle, na.rm = TRUE) /
              sum(events %in% at_bats_vector, na.rm = TRUE),
            at_bats = sum(events %in% at_bats_vector, na.rm = TRUE)) %>%
  mutate(xba = if_else(at_bats < 25, NA, xba)) %>%
  select(-at_bats)

# By side
xba_by_side <- data %>%
  group_by(player_name, p_throws) %>%
  summarize(xba = sum(estimated_ba_using_speedangle, na.rm = TRUE) /
              sum(events %in% at_bats_vector, na.rm = TRUE),
            at_bats = sum(events %in% at_bats_vector, na.rm = TRUE)) %>%
  mutate(xba = if_else(at_bats < 25, NA, xba)) %>%
  mutate(combined_name = paste("xba", p_throws, sep = "_")) %>%
  select(-p_throws, -at_bats) %>%
  pivot_wider(names_from = combined_name, values_from = xba)

# Combine
batter_xba <- batter_xba %>%
  left_join(xba_by_side, by = "player_name")

  
```

xBA last 25 PAs

```{r}
batter_xba_25 <- data %>%
  group_by(player_name) %>%
  arrange(desc(game_pk), at_bat_number) %>%
  mutate(pa_number = cumsum(events %in% plate_appearances_vector)) %>%
  summarize(xba_25 = sum(estimated_ba_using_speedangle *
                           (pa_number >= (max(pa_number) - 25)),
                         na.rm = TRUE) /
              sum(events %in% at_bats_vector *
                           (pa_number >= (max(pa_number) - 25)),
                          na.rm = TRUE))
```

PA per game

```{r}
batter_pa_per_game <- data %>%
  group_by(player_name) %>%
  summarize(pa_per_game = sum(events %in% plate_appearances_vector, na.rm = T) /
              length(unique(game_pk)))
```

ZO Swing %

```{r}
batter_zoswing <- data %>%
  group_by(player_name, p_throws, pitch_name) %>%
  summarize(pitches = n(),
            z_swing = sum(description %in% swings_vector & true_zone == 'strike', na.rm = T) /
                      sum(true_zone == 'strike', na.rm = T),
            o_swing = sum(description %in% swings_vector & true_zone == 'ball', na.rm = T) /
                      sum(true_zone == 'ball', na.rm = T),
            zo_swing = z_swing - o_swing) %>%
  mutate(zo_swing = if_else(pitches < 25, NA, zo_swing)) %>%
  select(-z_swing, -o_swing)%>%
  mutate(combined_name = paste(pitch_name, p_throws, "ZO", sep = "_")) %>%
  ungroup() %>%
  select(-p_throws, -pitches, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = zo_swing)
  


  
```

Join

```{r}
batter_stats <- batter_90 %>%
  left_join(batter_xba, by = "player_name") %>%
  left_join(batter_xba_25, by = "player_name") %>%
  left_join(batter_pa_per_game, by = "player_name") %>%
  left_join(batter_zoswing, by = "player_name") %>%
  mutate(date = '2023-07-04') %>%
  select(player_name, date, everything()) %>%
  # Remove sparse columns
  select(-c("Knuckleball_R_90", "Screwball_R_90", "Slurve_R_90",
            "Eephus_R_90", "Forkball_R_90", "Other_R_90", "Pitch Out_L_90", 
            "Pitch Out_R_90", "Other_L_90", "_L_90", "_R_90", "Knuckleball_R_ZO",
            "Screwball_R_ZO", "Eephus_R_ZO", "Forkball_R_ZO", "Other_R_ZO", 
            "Pitch Out_L_ZO", "Pitch Out_R_ZO", "Other_L_ZO", "_L_ZO","_R_ZO"))
```

#### Pitchers

90+%

```{r}
pitcher_90 <- data %>%
  group_by(pitcher, p_throws, pitch_name) %>%
  summarize(pitch_90 = sum(launch_speed >= 90, na.rm = TRUE) /
                            sum(description %in% swings_vector, na.rm = TRUE),
            swings = sum(description %in% swings_vector, na.rm = TRUE)) %>%
  mutate(pitch_90 = if_else(swings < 10, NA, pitch_90)) %>%
  mutate(combined_name = paste(pitch_name, p_throws, "90", "pitcher", sep = "_")) %>%
  ungroup() %>%
  select(-p_throws, -swings, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = pitch_90)
```

Usage

```{r}
pitcher_usage <- data %>%
  group_by(pitcher, p_throws, pitch_name) %>%
  summarize(pitches = n()) %>%
  group_by(pitcher) %>%
  mutate(total_pitches = sum(pitches),
         pitch_percentage = pitches / total_pitches,
         combined_name = paste(pitch_name, p_throws, "usage", sep = "_")) %>%
  select(-pitches, -total_pitches, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = pitch_percentage)

```

xBA last 75 BF

```{r}
pitcher_xba_75 <- data %>%
  group_by(pitcher) %>%
  arrange(desc(game_pk), at_bat_number) %>%
  mutate(pa_number = cumsum(events %in% plate_appearances_vector)) %>%
  summarize(xba_75_pitcher = sum(estimated_ba_using_speedangle *
                           (pa_number >= (max(pa_number) - 75)),
                         na.rm = TRUE) /
              sum(events %in% at_bats_vector *
                           (pa_number >= (max(pa_number) - 75)),
                          na.rm = TRUE))
```

BF per game

```{r}
pitcher_bf_per_game <- data %>%
  group_by(pitcher) %>%
  summarize(bf_per_game_pitcher = sum(events %in% plate_appearances_vector, na.rm = T) /
              length(unique(game_pk)))
```

#### Bullpen

```{r}
statcast_2023_1 <- statcast_2023 %>%
  left_join(starters, by='game_pk') %>%
   mutate(is_starter = case_when(
                                 pitcher == home_starter | pitcher == away_starter ~ 1,
                                 TRUE ~ 0))

data_1 <- statcast_2023_1 %>%
  filter(is_starter == 0 & game_date < '2023-07-04')

bullpen_xba_150 <- data_1 %>%
  group_by(pitcher_team) %>%
  arrange(desc(game_pk), at_bat_number) %>%
  mutate(pa_number = cumsum(events %in% plate_appearances_vector)) %>%
  summarize(xba_150_bullpen = sum(estimated_ba_using_speedangle *
                           (pa_number >= (max(pa_number) - 150)),
                         na.rm = TRUE) /
              sum(events %in% at_bats_vector *
                           (pa_number >= (max(pa_number) - 150)),
                          na.rm = TRUE))

```

## Combine into one loop to iterate over each day.

########################################################################

```{r message=FALSE, warning=FALSE}

all_dates <- unique(statcast_full$game_date)
first_date_for_row <- "2020-04-30" # do not create any rows for the first month of the season
row_dates <- all_dates[all_dates > first_date_for_row] 

full_batter_stats <- data.frame()
full_pitcher_stats <- data.frame()
full_bullpen_stats <- data.frame()

for(current_date in row_dates) {

data <- statcast_full %>%
  filter(game_date < current_date)

##################### Batter Stats ############################################

################ 90+

## By pitch & side
by_pitch_90 <- data %>%
  group_by(batter, p_throws, pitch_name) %>%
  summarize(pitch_90 = sum(launch_speed >= 90, na.rm = TRUE) /
                            sum(description %in% swings_vector, na.rm = TRUE),
            swings = sum(description %in% swings_vector, na.rm = TRUE)) %>%
  mutate(pitch_90 = if_else(swings < 10, NA, pitch_90))

batter_90 <- by_pitch_90 %>%
  mutate(combined_name = paste(pitch_name, p_throws, "90", sep = "_")) %>%
  ungroup() %>%
  select(-p_throws, -swings, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = pitch_90)

############ xBA

batter_xba <- data %>%
  group_by(batter) %>%
  summarize(xba = sum(estimated_ba_using_speedangle, na.rm = TRUE) /
              sum(events %in% at_bats_vector, na.rm = TRUE),
            at_bats = sum(events %in% at_bats_vector, na.rm = TRUE)) %>%
  mutate(xba = if_else(at_bats < 25, NA, xba)) %>%
  select(-at_bats)

# By side
xba_by_side <- data %>%
  group_by(batter, p_throws) %>%
  summarize(xba = sum(estimated_ba_using_speedangle, na.rm = TRUE) /
              sum(events %in% at_bats_vector, na.rm = TRUE),
            at_bats = sum(events %in% at_bats_vector, na.rm = TRUE)) %>%
  mutate(xba = if_else(at_bats < 25, NA, xba)) %>%
  mutate(combined_name = paste("xba", p_throws, sep = "_")) %>%
  select(-p_throws, -at_bats) %>%
  pivot_wider(names_from = combined_name, values_from = xba)

# Combine
batter_xba <- batter_xba %>%
  left_join(xba_by_side, by = "batter")

########## xBA 25

batter_xba_25 <- data %>%
  group_by(batter) %>%
  arrange(desc(game_pk), at_bat_number) %>%
  mutate(pa_number = cumsum(events %in% plate_appearances_vector)) %>%
  summarize(xba_25 = sum(estimated_ba_using_speedangle *
                           (pa_number >= (max(pa_number) - 25)),
                         na.rm = TRUE) /
              sum(events %in% at_bats_vector *
                           (pa_number >= (max(pa_number) - 25)),
                          na.rm = TRUE))

########### PA per game

batter_pa_per_game <- data %>%
  group_by(batter) %>%
  summarize(pa_per_game = sum(events %in% plate_appearances_vector, na.rm = T) /
              length(unique(game_pk)))

########### ZO Swing

batter_zoswing <- data %>%
  group_by(batter, p_throws, pitch_name) %>%
  summarize(pitches = n(),
            z_swing = sum(description %in% swings_vector & true_zone == 'strike', na.rm = T) /
                      sum(true_zone == 'strike', na.rm = T),
            o_swing = sum(description %in% swings_vector & true_zone == 'ball', na.rm = T) /
                      sum(true_zone == 'ball', na.rm = T),
            zo_swing = z_swing - o_swing) %>%
  mutate(zo_swing = if_else(pitches < 25, NA, zo_swing)) %>%
  select(-z_swing, -o_swing)%>%
  mutate(combined_name = paste(pitch_name, p_throws, "ZO", sep = "_")) %>%
  ungroup() %>%
  select(-p_throws, -pitches, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = zo_swing)

############### Join

batter_stats <- batter_90 %>%
  left_join(batter_xba, by = "batter") %>%
  left_join(batter_xba_25, by = "batter") %>%
  left_join(batter_pa_per_game, by = "batter") %>%
  left_join(batter_zoswing, by = "batter") %>%
  mutate(date = current_date) %>%
  select(batter, date, everything())


############################## Pitcher Stats ###################################

############# 90 + 

pitcher_90 <- data %>%
  group_by(pitcher, p_throws, pitch_name) %>%
  summarize(pitch_90 = sum(launch_speed >= 90, na.rm = TRUE) /
                            sum(description %in% swings_vector, na.rm = TRUE),
            swings = sum(description %in% swings_vector, na.rm = TRUE)) %>%
  mutate(pitch_90 = if_else(swings < 10, NA, pitch_90)) %>%
  mutate(combined_name = paste(pitch_name, p_throws, "90", "pitcher", sep = "_")) %>%
  ungroup() %>%
  select(-p_throws, -swings, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = pitch_90)

############# Usage

pitcher_usage <- data %>%
  group_by(pitcher, p_throws, pitch_name) %>%
  summarize(pitches = n()) %>%
  group_by(pitcher) %>%
  mutate(total_pitches = sum(pitches),
         pitch_percentage = pitches / total_pitches,
         combined_name = paste(pitch_name, p_throws, "usage", sep = "_")) %>%
  select(-pitches, -total_pitches, -pitch_name) %>%
  pivot_wider(names_from = combined_name, values_from = pitch_percentage) %>%
  # change all NA to 0% usage
  mutate(across(-p_throws, ~if_else(is.na(.), 0, .))) # ~ means "function"


############# xBA last 75

pitcher_xba_75 <- data %>%
  group_by(pitcher) %>%
  arrange(desc(game_pk), at_bat_number) %>%
  mutate(pa_number = cumsum(events %in% plate_appearances_vector)) %>%
  summarize(xba_75_pitcher = sum(estimated_ba_using_speedangle *
                           (pa_number >= (max(pa_number) - 75)),
                         na.rm = TRUE) /
              sum(events %in% at_bats_vector *
                           (pa_number >= (max(pa_number) - 75)),
                          na.rm = TRUE))

############## BF per game

pitcher_bf_per_game <- data %>%
  group_by(pitcher) %>%
  summarize(bf_per_game_pitcher = sum(events %in% plate_appearances_vector, na.rm = T) /
              length(unique(game_pk)))

############## Join

pitcher_stats <- pitcher_90 %>%
  left_join(pitcher_usage, by = 'pitcher') %>%
  left_join(pitcher_xba_75, by = 'pitcher') %>%
  left_join(pitcher_bf_per_game, by = 'pitcher') %>%
  mutate(date = current_date) %>%
   select(pitcher, date, everything())

#################################### Bullpen Stat #############################

########### xba 150

statcast_full_1 <- statcast_full %>%
  left_join(starters, by='game_pk') %>%
   mutate(is_starter = case_when(
                                 pitcher == home_starter | pitcher == away_starter ~ 1,
                                 TRUE ~ 0))

data_1 <- statcast_full_1 %>%
  filter(is_starter == 0 & game_date < current_date)

bullpen_stats <- data_1 %>%
  group_by(pitcher_team) %>%
  arrange(desc(game_pk), at_bat_number) %>%
  mutate(pa_number = cumsum(events %in% plate_appearances_vector)) %>%
  summarize(xba_150_bullpen = sum(estimated_ba_using_speedangle *
                           (pa_number >= (max(pa_number) - 150)),
                         na.rm = TRUE) /
              sum(events %in% at_bats_vector *
                           (pa_number >= (max(pa_number) - 150)),
                          na.rm = TRUE)) %>%
  mutate(date = current_date)


########################################################################

full_batter_stats <- bind_rows(full_batter_stats, batter_stats)
full_pitcher_stats <- bind_rows(full_pitcher_stats, pitcher_stats)
full_bullpen_stats <- bind_rows(full_bullpen_stats, bullpen_stats)


}

```


```{r}
summary(batter_stats)
```

```{r}
summary(bullpen_stats)
```

```{r}
summary(game_data)
```

```{r}

full_data <- game_data %>%
  left_join(full_batter_stats, by = c('batter', 'date')) %>%
  left_join(full_pitcher_stats, by = c('opp_starter' = 'pitcher', 'date')) %>%
  left_join(full_bullpen_stats, by = c('pitcher_team', 'date')) %>%
  select(date, game_pk, batter, batter_team, opp_starter, pitcher_team, hits, everything()) 

```


```{r}
head(full_data)
```








